---
author: Peter Boshe
date: '2021-09-30'
title: Popular Polutants of the Public `r emo::ji("litter")
output: html_document 
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', collapse = TRUE, tidy = 'styler')
```


The data was collected by [Wastebase's Data Partner](https://www.unwaste.io/partners) by methods decribed on this [source page](https://www.kaggle.com/wastebase/plastic-bottle-waste)

## setting the environment

```{r data, warning=FALSE, message=FALSE}

rm(list = ls())
require(tidyverse)
require(glue)
require(knitr)
require(DT)
require(RColorBrewer)
require(wordcloud2)

palette <- brewer.pal(n = 4,
                      name = "RdYlBu")

setwd("~/Portfolio/web-tea/content/post/2021-09-30-popular-pollutants-of-the-public/index_files")
files <- list.files(pattern = "*F.csv")
all_files <- map(files, read_csv)

names(all_files) <- month.abb[3:12] #to name the data according to month created which in our case is march to december

combined_file <- bind_rows(all_files, .id = "batch")
```


## pre-processing & cleaning our dataset

```{r cleaning}

df <- combined_file %>%
#remove the bottle weight column
#remove the url column
  select(-bottle_weight, -data_url, -product_label, -product_barcode)

#convert batch, manufacturer_name, manufacturer_country, scan_country into factors, months
df$batch <- as.factor(df$batch)
df$manufacturer_country <- as.factor(df$manufacturer_country)
df$manufacturer_name <- as.factor(df$manufacturer_name)
df$scan_country <- as.factor(df$scan_country)
df$brand_name <- as.factor(df$brand_name)

#parse product_size for uniform measurements
number_pattern <- "^\\d\\s?\\.?\\d*"
unit_pattern <- "\\D+$"

df1 <- df %>%
  mutate(product_size_extracted = str_extract(product_size, number_pattern)) %>%
  mutate(units = str_extract(product_size, unit_pattern)) %>%
  drop_na(units)

df1$units <- str_trim(df1$units)
df1$product_size_extracted <- as.numeric(str_trim(df1$product_size_extracted))

df1$units <- ifelse(df1$units == "litre", "l", df1$units)
df1$units <- ifelse(df1$units == "ltrs", "l", df1$units)
df1$units <- ifelse(df1$units == "ltr", "l", df1$units)
df1$units <- ifelse(df1$units == "Litres", "l", df1$units)
df1$units <- ifelse(df1$units == "L", "l", df1$units)
df1$units <- ifelse(df1$units == "gm", "g", df1$units)
df1$units <- ifelse(df1$units == "m", "ml", df1$units)


df2 <- df1 %>%
  mutate(
    amount = case_when(
      units == "l" ~ product_size_extracted * 1000,
      units == "kg" ~ product_size_extracted * 1000,
      TRUE ~ product_size_extracted
    )
    )

df2$units <- ifelse(df2$units == "l", "ml", df2$units)
df2$units <- ifelse(df2$units == "kg", "g", df2$units)

data <- df2 %>%
  select(batch,
         brand_name,
         manufacturer_name,
         manufacturer_country,
         scan_country,
         amount,
         units,
         bottle_count) %>%
 mutate(units = as.factor(units))

# we are left with columns
glue("We are left with the columns; {columns}",
     columns = glue_collapse(colnames(data),
                             sep = ", ",
                             last = ", and ")
)

```

## clean our environment

```{r housekeeping}

rm(list = c("all_files",
            "combined_file",
            "df",
            "df1",
            "df2",
            "number_pattern",
            "unit_pattern"))


```

### Total bottles per country

```{r bottles_per_country}

data |> 
  group_by(scan_country) |> 
  summarise(total_bottles = sum(bottle_count)) |> 
  arrange(desc(total_bottles))

```

 + From the entire dataset we see that my beautiful country comes in second while
 + Mozambique holds the lead with more than 200,000 bottles scanned.

### Table of the top 20 manufacterers

```{r bottles_per_manufacterer}

tz_data <- data %>%
        filter(scan_country == "TZ")   #filter data for Tanzania (case study)


tz_manufacturer_country  <-  tz_data %>%
  group_by(manufacturer_name, manufacturer_country) %>%
  summarise(total_bottles = sum(bottle_count)) %>%
  arrange(desc(total_bottles)) %>% 
  head(20)


datatable(tz_manufacturer_country)

```

+ From the table above we can see that it is a two-horse race on the beverage heavy weights, METL and BAHKHRESA,
lets visualize the contrast between the numbers


### Chart of the top 20 manufacterers

```{r distribution}

tz_manufacturer_country %>% 
  ggplot(aes(x = total_bottles, 
             y = reorder(manufacturer_name,total_bottles), 
             color = total_bottles)) +
  geom_segment(aes(yend = manufacturer_name), 
               xend = 0,
               size = 2) + 
  geom_point(size = 9) +   # shape = 21) +
  geom_text(aes(label = total_bottles), 
            color = "black",
            size = 2.5) + 
  scale_color_gradientn(colors = palette, 
                        trans = "reverse") +
  geom_curve(aes(x = 2000, 
                 y = 5,
                 xend = 4000,
                 yend = 8), 
             curvature = 0.2,
             angle = 45, 
             ncp = 5,
             color = "black",
             arrow = arrow(angle = 30, 
                           length = unit(0.5, "cm"), 
                           ends = "first", 
                           type = "open"))+
  geom_vline(xintercept = mean(tz_manufacturer_country$total_bottles), 
             linetype="dotdash",
             color = "#4E7FBF", 
             size = 0.5) +
  theme_minimal() +
  theme(
    axis.line.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_blank(),
    legend.position="none",
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.line.x = element_blank(), 
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank())  +
  labs(title = "20 Manufacturers with the most litter collected, 2021",
       caption = "Source: Wastebase.org") +
  xlab("total quantity of bottles collected") +
  ylab("") +
  annotate(
    "text",
    family = "sans",
    fontface = "italic",
    x = 5000, 
    y = 10, 
    label = "Average amount\n of litter collected\n per manufacturer"
  )

```

### Top 10 reoccurring brands


```{r top_ten}
# here it can be observed that the 
tz_data %>%
  group_by(manufacturer_country) %>%
  summarise(total_bottles = sum(bottle_count)) %>%
  arrange(desc(total_bottles)) %>% 
  head(10)

tz_data %>% 
  group_by(brand_name) %>% 
  summarise(total = sum(bottle_count)) %>% 
  arrange(desc(total))


```

### distribution of container size

```{r container_size_distribution}

tz_data %>% 
  filter(units == "ml") %>% 
  ggplot(aes(amount)) +
  geom_histogram(fill = "#4E7FBF",
                 binwidth = 5, 
                 breaks = (seq(-250,15000, by = 500))) + 
  geom_curve(aes(x = 500, 
                 y = 730,
                 xend = 5000,
                 yend = 600), 
             curvature = -0.4,
             angle = 45, 
             ncp = 5, 
             arrow = arrow(angle = 30, 
                           length = unit(0.5, "cm"), 
                           ends = "first", 
                           type = "open")) +
  geom_curve(aes(x = 13000, 
                 y = 12,
                 xend = 10000,
                 yend = 300), 
             curvature = 0.2,
             angle = 45, 
             ncp = 5, 
             arrow = arrow(angle = 30, 
                           length = unit(0.5, "cm"), 
                           ends = "first", 
                           type = "open")) +
  geom_curve(aes(x = 6000, 
                 y = 12,
                 xend = 10000,
                 yend = 300), 
             curvature = -0.2,
             angle = 45, 
             ncp = 5, 
             arrow = arrow(angle = 30, 
                           length = unit(0.5, "cm"), 
                           ends = "first", 
                           type = "open")) +
  theme_minimal() +
  theme(
    panel.grid.minor.y = element_blank(),
    # panel.grid.major.y = element_blank(),
    legend.position="none", 
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text = element_text(color = "black"))  +
  xlab("waste container capacity (ml)") +
  ylab("") + 
  annotate(
    "text",
    family = "sans",
    fontface = "italic",
    x = 5000, 
    y = 550, 
    label = "Containers\n of \n 250-500ml"
  ) + 
  annotate(
    "text",
    family = "sans",
    fontface = "italic",
    x = 10000, 
    y = 360, 
    label = "Containers\n of \n 6 & 13 liters"
  )

```

### creating word cloud with brand names

```{r word_cloud}

mywords <- tz_data %>% 
  group_by(brand_name) %>% 
  count() %>% 
  arrange(desc(n))

wordcloud2(mywords, 
           minRotation = -pi/16, 
           maxRotation = -pi/16,
           backgroundColor = "white", 
           color="random-dark", 
           shape = 'triangle', 
           ellipticity = 0.9)


```


+ inspecting the general trends for the bottle count per nation

+ relevelling the batch column


### line chart for monthly trends per country

```{r trend_per_nation}

data$batch <- factor(data$batch, 
                     levels = c("Mar", 
                                "Apr", 
                                "May", 
                                "Jun", 
                                "Jul", 
                                "Aug",
                                "Sep",
                                "Oct",
                                "Nov",
                                "Dec"))



data %>% 
  complete(scan_country, batch) %>% 
  group_by(scan_country, batch) %>% 
  summarise(total_bottles = sum(bottle_count)) %>% 
  replace_na(list(total_bottles = 0)) %>%
  ggplot(aes(batch, total_bottles, color = scan_country)) +
  geom_line(aes(group = scan_country))

```




































