---
title: "Visualization - Case Study Selection"
description: "Visualization project to mine, analyse and present data to justify selection criteria for a case study"
author: "Peter Boshe"
date: 2021-05-01T11:00:59-04:00
categories: ["R"]
tags: ["R programming","Visualization","Data Analysis"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

The task was to mine for relevant data and presenting them in a manner that would simplify the decision making process
in picking countries as case studies to assess the state and level of data protection regulation in Africa in order to;
- Get a deeper feel of the trends present among the countries; to find out which nations are to a more degree exposed to data mishandling;
- To find out which countries have strategies in place to safeguard their citizens' right of privacy,

>The following are some of the visuals that I had then created to facilitate the research paper.


```{r cars}
summary(cars)
fit <- lm(dist ~ speed, data = cars)
fit
```

# Including Plots

You can also embed plots. See Figure \@ref(fig:pie) for example:

```{r pie, fig.cap='A fancy pie chart.', tidy=FALSE}
par(mar = c(0, 1, 0, 1))
pie(
  c(280, 60, 20),
  c('Sky', 'Sunny side of pyramid', 'Shady side of pyramid'),
  col = c('#0292D8', '#F7EA39', '#C4B632'),
  init.angle = -50, border = NA
)
```
