---
title: Debt in the East; 1970 - 2020 `r emo::ji("burundi")` `r emo::ji("kenya")` `r emo::ji("rwanda")` `r emo::ji("tanzania")` `r emo::ji("uganda")`
date: 2022-01-25T19:49:17+03:00
author: "Peter Boshe"
categories: ["Socio-eco Welfare"]
tags: ["DataViz", "Unguided project", "E.D.A"]
output:
  blogdown::html_page:
    toc: true
    toc_depth: 5
images: []
summary: "In this Article we explore the trends of debt and major events in Burundi, Kenya, Rwanda, Tanzania,  and Uganda."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', collapse = TRUE, echo = FALSE, message = FALSE)
```

```{r}
# clear the environment
rm(list = ls())
# set up the environment
library(tidyverse)
library(lubridate)
library(DT)
library(janitor)
library(naniar)
library(ggthemes)
library(ggdark)
library(tidylog)
# load our dataset
# data <- read_csv("/home/pedro/Portfolio/web-tea/content/post/debt_in_the_east/data.csv")
# metadata <- read_csv("/home/pedro/Portfolio/web-tea/content/post/debt_in_the_east/metadata.csv")
data2 <- read_csv("/home/pedro/Portfolio/web-tea/content/post/debt_in_the_east/data2.csv")
metadata2 <- read_csv("/home/pedro/Portfolio/web-tea/content/post/debt_in_the_east/metadata2.csv")
# data <- clean_names(data)
data2 <- clean_names(data2)
```

---

## Our Data

Our data was downloaded from the [international debt statistics](https://databank.worldbank.org/source/international-debt-statistics#) which was last updated on the 14th of January at the time of download.

We have data on five countries in East Africa to Explore and compare some debt metrics from 1970-2020. The five countries are;

-   Burundi
-   Kenya
-   Uganda
-   Rwanda
-   Tanzania

Note: This Analysis is purely exploratory in nature, and correlation is not proof of causation.

---

### Current Account Balance (current US$)

This introductory plot shows the trends of the `current account balance` and some major events that took place during our time period of interest. Although there might be several first-takes but mine was, **"what happened in the 2000s ? and is it still happening?"**

This plot, however, does not give granular insight to trends. As one would wonder why country's account balances took steep dives even after events like [SAPs transitions](https://www.imf.org/external/np/prsp/prsp.aspx), [G8 summit in 2005](https://en.wikipedia.org/wiki/31st_G8_summit) and the [2000 Jubilee.](https://en.wikipedia.org/wiki/Jubilee_2000)

The `current account balance` is a result of various factors like new grants, disbursements, debt forgiveness grants, principal repayments to external debt, reserves and many more.

```{r}
## specifying our focus groups

nations <- c("Tanzania", "Uganda", "Kenya", "Rwanda", "Burundi")

princ_other_series <- c("Principal forgiven (current US$)", 
                        "Principal rescheduled (current US$)", 
                        "Principal forgiven (current US$)", 
                        "Principal rescheduled, official (current US$)",
                        "Principal rescheduled, private (current US$)")
cur_series <- c("Present value of external debt (current US$)",
                "Debt buyback (current US$)", 
                "Current account balance (current US$)", 
                "Debt forgiveness grants (current US$)", 
                "Debt forgiveness or reduction (current US$)",
                "Technical cooperation grants (current US$)",
                "Total amount of debt rescheduled (current US$)",
                "Total reserves (includes gold, current US$)")
  
ppg_comp_series <- str_subset(unique(data2$series_name), '^Currency composition.*')
disburse_series <- str_subset(unique(data2$series_name), '.*isburse.*')
netflow_series <- str_subset(unique(data2$series_name), '^Net.*')
princ_repay_series <- str_subset(unique(data2$series_name), '^Principal repayments on external debt.*')
imf_cred_series <- str_subset(unique(data2$series_name), '^Use of IMF credit.*')
debt_service <- str_subset(unique(data2$series_name), '^Debt service.*')
central_bank_series <- str_subset(unique(data2$series_name), '.*central bank.*')
government_series <- str_subset(unique(data2$series_name), '.*government sector.*')
```

```{r}
## Recoding data2$series_name into data2$series_name_rec
data2$series_name_rec <- data2$series_name %>%
  fct_recode(
    "Current account balance" = "Current account balance (current US$)",
    "Debt buyback" = "Debt buyback (current US$)",
    "Debt forgiveness grants" = "Debt forgiveness grants (current US$)",
    "Debt forgiveness or reduction" = "Debt forgiveness or reduction (current US$)",
    "Debt service on external debt, central bank (PPG)" = "Debt service on external debt, central bank (PPG) (TDS, current US$)",
    "Debt service on external debt, general government sector (PPG)" = "Debt service on external debt, general government sector (PPG) (TDS, current US$)",
    "Debt service on external debt, long-term" = "Debt service on external debt, long-term (TDS, current US$)",
    "Debt service on external debt, other public sector (PPG)" = "Debt service on external debt, other public sector (PPG) (TDS, current US$)",
    "Debt service on external debt, private guaranteed by public sector (PPG)" = "Debt service on external debt, private guaranteed by public sector (PPG) (TDS, current US$)",
    "Debt service on external debt, private nonguaranteed (PNG)" = "Debt service on external debt, private nonguaranteed (PNG) (TDS, current US$)",
    "Debt service on external debt, public and publicly guaranteed (PPG)" = "Debt service on external debt, public and publicly guaranteed (PPG) (TDS, current US$)",
    "Debt service on external debt, public sector (PPG)" = "Debt service on external debt, public sector (PPG) (TDS, current US$)",
    "Debt service on external debt, total" = "Debt service on external debt, total (TDS, current US$)",
    "Disbursements on external debt, central bank (PPG)" = "Disbursements on external debt, central bank (PPG) (DIS, current US$)",
    "Disbursements on external debt, general government sector (PPG)" = "Disbursements on external debt, general government sector (PPG) (DIS, current US$)",
    "Disbursements on external debt, long-term" = "Disbursements on external debt, long-term (DIS, current US$)",
    "Disbursements on external debt, long-term + IMF" = "Disbursements on external debt, long-term + IMF (DIS, current US$)",
    "Disbursements on external debt, other public sector (PPG)" = "Disbursements on external debt, other public sector (PPG) (DIS, current US$)",
    "Disbursements on external debt, private guaranteed by public sector (PPG)" = "Disbursements on external debt, private guaranteed by public sector (PPG) (DIS, current US$)",
    "Disbursements on external debt, private nonguaranteed (PNG)" = "Disbursements on external debt, private nonguaranteed (PNG) (DIS, current US$)",
    "Disbursements on external debt, public and publicly guaranteed (PPG)" = "Disbursements on external debt, public and publicly guaranteed (PPG) (DIS, current US$)",
    "Disbursements on external debt, public sector (PPG)" = "Disbursements on external debt, public sector (PPG) (DIS, current US$)",
    "Net flows on external debt, central bank (PPG)" = "Net flows on external debt, central bank (PPG) (NFL, current US$)",
    "Net flows on external debt, general government sector (PPG)" = "Net flows on external debt, general government sector (PPG) (NFL, current US$)",
    "Net flows on external debt, long-term" = "Net flows on external debt, long-term (NFL, current US$)",
    "Net flows on external debt, other public sector (PPG)" = "Net flows on external debt, other public sector (PPG) (NFL, current US$)",
    "Net flows on external debt, private guaranteed by public sector (PPG)" = "Net flows on external debt, private guaranteed by public sector (PPG) (NFL, current US$)",
    "Net flows on external debt, private nonguaranteed (PNG)" = "Net flows on external debt, private nonguaranteed (PNG) (NFL, current US$)",
    "Net flows on external debt, public and publicly guaranteed (PPG)" = "Net flows on external debt, public and publicly guaranteed (PPG) (NFL, current US$)",
    "Net flows on external debt, public sector (PPG)" = "Net flows on external debt, public sector (PPG) (NFL, current US$)",
    "Net flows on external debt, short-term" = "Net flows on external debt, short-term (NFL, current US$)",
    "Net flows on external debt, total" = "Net flows on external debt, total (NFL, current US$)",
    "Present value of external debt" = "Present value of external debt (current US$)",
    "Principal forgiven" = "Principal forgiven (current US$)",
    "Principal repayments on external debt, central bank (PPG)" = "Principal repayments on external debt, central bank (PPG) (AMT, current US$)",
    "Principal repayments on external debt, general government sector (PPG)" = "Principal repayments on external debt, general government sector (PPG) (AMT, current US$)",
    "Principal repayments on external debt, long-term" = "Principal repayments on external debt, long-term (AMT, current US$)",
    "Principal repayments on external debt, long-term + IMF" = "Principal repayments on external debt, long-term + IMF (AMT, current US$)",
    "Principal repayments on external debt, other public sector (PPG)" = "Principal repayments on external debt, other public sector (PPG) (AMT, current US$)",
    "Principal repayments on external debt, private guaranteed by public sector (PPG)" = "Principal repayments on external debt, private guaranteed by public sector (PPG) (AMT, current US$)",
    "Principal repayments on external debt, private nonguaranteed (PNG)" = "Principal repayments on external debt, private nonguaranteed (PNG) (AMT, current US$)",
    "Principal repayments on external debt, public and publicly guaranteed (PPG)" = "Principal repayments on external debt, public and publicly guaranteed (PPG) (AMT, current US$)",
    "Principal repayments on external debt, public sector (PPG)" = "Principal repayments on external debt, public sector (PPG) (AMT, current US$)",
    "Principal rescheduled" = "Principal rescheduled (current US$)",
    "Principal rescheduled, official" = "Principal rescheduled, official (current US$)",
    "Principal rescheduled, private" = "Principal rescheduled, private (current US$)",
    "Technical cooperation grants" = "Technical cooperation grants (current US$)",
    "Total amount of debt rescheduled" = "Total amount of debt rescheduled (current US$)",
    "Total change in external debt stocks" = "Total change in external debt stocks (current US$)",
    "Undisbursed external debt, official creditors" = "Undisbursed external debt, official creditors (UND, current US$)",
    "Undisbursed external debt, private creditors" = "Undisbursed external debt, private creditors (UND, current US$)",
    "Undisbursed external debt, total" = "Undisbursed external debt, total (UND, current US$)"
  )
```

```{r}
### Create a dataframe of events
events <- data.frame(year = c(1977, 1978, 1980, 1985, 1991, 1992, 1996, 1998, 2002, 2005, 2007, 2011, 2019, 2020),
                     event = c("EAC Break-up", "Kagera War (1978 - 1979)", "\n1st general elections\nafter Idi-amin(UG)", "End of Ujamaa(TZ)", "KANU one party rule ends", "Multiparty Democracy in TZ", "\nHIPC Initiative (KNY excl.)", "\n2nd Congo War\n(1998 - 2003)", "SAPs transition", "G8 gathering for debt relief", "\nGlobal Financial crisis\n(2007 - 2008)", "\nEast Africa Drought\n(2011 - 2012)", "Covid-19 Outbreak", "Central & East Africa Floods"))

events <- mutate(events, year = as.character(events$year))


df1 <- data2 %>%
    filter(country_name %in% nations) %>%
    filter(series_name %in% cur_series) %>%
    group_by(series_name) %>%
    pivot_longer(cols = contains("_yr"), names_to = "year") %>%
    mutate(year = str_match(year, "\\d{4}")) %>%
    select(
        year,
        country_name,
        series_name,
        series_name_rec,
        value
    ) %>%
    mutate(value = as.numeric(na_if(value, ".."))) %>%
    replace(is.na(.), 0) %>%
    mutate(value = value / 10^6) %>%
    clean_names() %>%
    filter(series_name %in% c("Current account balance (current US$)"))


df2 <- df1 %>% left_join(events, by = "year")

```

```{r, warning=FALSE, fig.height = 8, fig.width = 12.5, fig.align = "center"}


df2 %>%
  ggplot(aes(year, value, color = country_name)) +
  geom_line(aes(group = country_name)) +
  scale_x_discrete(
        breaks = seq(1970, 2020, 2), 
        position = "top") +
  theme_solarized_2() +
    theme(axis.text.x = element_text(angle = 90),
        line = element_blank(), 
        plot.background = element_rect(fill = "#F5E6B6"),
        plot.title = element_text(hjust = 0.5),
        legend.position = "bottom",
        legend.text = element_text(size = 13,
                                   color = "black")) +
  # scale_fill_manual(values=wes_palette(1, "FantasticFox1")) +
  # scale_color_brewer(palette="Dark2") +
    scale_color_manual(
    values = c(Kenya = "#FF0000",
    Rwanda = "#046C9A",
    Tanzania = "#00A08A",
    Uganda = "#000000",
    Burundi = "#F98400")
  ) +
  # scale_x_discrete(position = "top") +
    labs(
    title = "Current Account Balance (current US$)", 
    x = " ",
    y = "million USDs",
    caption = " data: World Bank, international debt statistics (1970-2020)\n https://peterboshe.netlify.app/",
    color = "COUNTRY"
  ) +
  geom_vline(data = events, aes(xintercept = year),
                                color = "grey",
                                linetype = "dashed") +
    geom_text(aes(x = year,
                y = -5500,
                label = event,
                angle = 90), 
                size = 4,
              nudge_x = 0.3,
              color = "darkgrey",
              fontface = "italic") +
  annotate("rect", xmin = "1978", xmax = "1979", ymin = -Inf, ymax = Inf,
              alpha = .2, fill = "grey") + 
  annotate("rect", xmin = "1998", xmax = "2003", ymin = -Inf, ymax = Inf,
              alpha = .2, fill = "grey") + 
  annotate("rect", xmin = "2007", xmax = "2008", ymin = -Inf, ymax = Inf,
              alpha = .2, fill = "grey") +
  annotate("rect", xmin = "2011", xmax = "2012", ymin = -Inf, ymax = Inf,
              alpha = .2, fill = "grey") 



```


---

In the following sections we will be looking at some of the factors that account to the `current account balance` sum, and we will see how they relate with some of the major events. Let's dive right in. 

---

#### 1. Debt forgiveness grants (current US$)

We see a collective spike in debt forgiveness grants after the G8 summit (so meetings do work!), however Kenya and Burundi seem to be left out on the action. It would be interesting to know if they were withheld due to not meeting the same criteria that left Kenya out of the HIPC initiative in 1996. 
(links to resources below plot)


```{r}
data2 %>%
    filter(country_name %in% nations) %>%
    filter(series_name %in% cur_series) %>%
    group_by(series_name) %>%
    pivot_longer(cols = contains("_yr"), names_to = "year") %>%
    mutate(year = str_match(year, "\\d{4}")) %>%
    select(
        year,
        country_name,
        series_name,
        series_name_rec,
        value
    ) %>%
    mutate(value = as.numeric(na_if(value, ".."))) %>%
    replace(is.na(.), 0) %>%
    mutate(value = value / 10^6) %>%
    clean_names() %>%
    filter(series_name %in% c("Debt forgiveness grants (current US$)")) %>%
  left_join(events, by = "year") %>%
     ggplot(aes(year, value, color = country_name)) +
  geom_line(aes(group = country_name)) +
   geom_curve(aes(x = "2004", 
                 y = 2500,
                 xend = "1998",
                 yend = 2500),
              color = "darkgrey",
             curvature = 0.2,
             angle = 45, 
             ncp = 5, 
             arrow = arrow(angle = 30, 
                           length = unit(0.5, "cm"), 
                           ends = "first", 
                           type = "open")) +
  annotate("text", x = "1995", y = 2500, label = "G8 summit", size = 3, color = "black", fontface = "italic") +
  scale_x_discrete(
        breaks = seq(1970, 2020, 2)) +
  theme(axis.text.x = element_text(angle = 90), 
        panel.background = element_rect(fill = "#F5E6B6")) +
    geom_vline(xintercept = "2005", 
               color = "grey", 
               linetype = "dashed") + 
  theme_hc() +
    scale_color_manual(
    values = c(Kenya = "#FF0000",
    Rwanda = "#046C9A",
   Tanzania = "#00A08A",
    Uganda = "#000000",
    Burundi = "#F98400")
  ) +
    labs(
    title = "Debt forgiveness grants (current US$)", 
    x = " ",
    y = "million USDs",
    caption = "- data from 1970-2020\n- each y-axis is independent",
    color = ""
  )
```


-   In June **2005**, the Group of Eight (G8) gathering of the worlds most developed economies, resolved, through their finance ministers, to grant total debt relief to 18 countries that had reached 'completion point', amounting to \$40 billion. Fourteen of these countries are in Africa. [Foreign aid, debt relief and Africa's development: problems and prospects. 2009.](https://www.tandfonline.com/doi/pdf/10.1080/10220460902986180)

-   The World Bank, the International Monetary Fund (IMF) and other multilateral, bilateral and commercial creditors began the Heavily Indebted Poor Countries (HIPC) Initiative in **1996**, However among the East African countries, Kenya does not meet the criteria to receive debt relief. Countries must make full use of traditional debt-relief mechanisms to be eligible for debt relief under the HIPC Initiative. [Brief](https://www.worldbank.org/en/topic/debt/brief/hipc) [Countries must make full use of traditional debt-relief mechanisms to be eligible for debt relief under the HIPC Initiative.](https://www.imf.org/external/np/hipc/cost2/index.htm)

---

#### 2. Total reserves (includes gold, current US$)

For some reason, As of 2022-02-16, The world bank's most recent info reported on Total reserves for both Tanzania and Uganda is from **2018**.




```{r}
data2 %>%
    filter(country_name %in% nations) %>%
    filter(series_name %in% cur_series) %>%
    group_by(series_name) %>%
    pivot_longer(cols = contains("_yr"), names_to = "year") %>%
    mutate(year = str_match(year, "\\d{4}")) %>%
    select(
        year,
        country_name,
        series_name,
        series_name_rec,
        value
    ) %>%
    mutate(value = as.numeric(na_if(value, ".."))) %>%
    replace(is.na(.), 0) %>%
    mutate(value = value / 10^6) %>%
    clean_names() %>%
    filter(series_name %in% c("Total reserves (includes gold, current US$)")) %>%
  left_join(events, by = "year") %>%
     ggplot(aes(year, value, color = country_name)) +
  geom_line(aes(group = country_name)) +
  geom_curve(aes(x = "2018", 
                 y = 8000,
                 xend = "1998",
                 yend = 8000),
              color = "darkgrey",
             curvature = 0.2,
             angle = 45, 
             ncp = 5, 
             arrow = arrow(angle = 30, 
                           length = unit(0.5, "cm"), 
                           ends = "first", 
                           type = "open")) +
  annotate("text", x = "1997", y = 8000, label = "??", size = 3, color = "black", fontface = "italic") +
  scale_x_discrete(
        breaks = seq(1970, 2020, 2)) +
  theme(axis.text.x = element_text(angle = 90), 
        panel.background = element_rect(fill = "#F5E6B6")) +
        geom_vline(xintercept = "2018", 
               color = "grey", 
                 linetype = "dashed") +
  theme_hc() +
    scale_color_manual(
    values = c(Kenya = "#FF0000",
    Rwanda = "#046C9A",
    Tanzania = "#00A08A",
    Uganda = "#000000",
    Burundi = "#F98400")
  ) +
    labs(
    title = "Total reserves (includes gold, current US$)", 
    x = " ",
    y = "million USDs",
    caption = "- data from 1970-2020\n- each y-axis is independent",
    color = "COUNTRY"
  )
```

---

#### 3. Debt service and Net Flows on external debt, central bank

It is curious to see how the trends of `Debt service on external debt, central bank` correlated with political unrest and wars with the respective countries involved.

The Ugandan Bush war intrigued me as even with other charts on `Disbursement on external debt` and `Principal payment on external debt` (which are not incl. due to redundancy) show the same trends, where the `payment of external debt, central bank`, peaked during the war. Even the `Netflow on external debt` took a hard regression towards the mean to negative values. Would be interesting to learn more on this.  
(links to resources below plot)

```{r, fig.height = 4, fig.width = 12.5, fig.align = "center"}
# we now wrangle our data

data2 %>%
  filter(country_name %in% nations) %>%
  filter(series_name %in% central_bank_series) %>%
  group_by(series_name) %>%
  pivot_longer(cols = contains("_yr"), names_to = "year") %>%
  mutate(year = str_match(year, "\\d{4}")) %>%
  select(
    year,
    country_name,
    series_name,
    series_name_rec,
    value
  ) %>%
  mutate(value = as.numeric(na_if(value, ".."))) %>%
  replace(is.na(.), 0) %>%
  mutate(value = value / 10^6) %>%
  clean_names() %>%
   filter(series_name %in% c("Debt service on external debt, central bank (PPG) (TDS, current US$)", "Net flows on external debt, central bank (PPG) (NFL, current US$)")) %>%
    ggplot(aes(year, value, color = country_name)) +
  geom_line(aes(group = country_name)) +
  scale_x_discrete(
        breaks = seq(1970, 2020, 2)) +
  theme(axis.text.x = element_text(angle = 90), 
        panel.background = element_rect(fill = "#F5E6B6")) +
  theme_hc() +
    scale_color_manual(
    values = c(Kenya = "#FF0000",
    Rwanda = "#046C9A",
    Tanzania = "#00A08A",
    Uganda = "#000000",
    Burundi = "#F98400")
  ) +
    labs(
    x = " ",
    y = "million USDs",
    caption = "- data from 1970-2020\n- each y-axis is independent",
    color = "COUNTRY"
  ) +
  annotate("rect", xmin = "1980", xmax = "1986", ymin = -Inf, ymax = Inf,
              alpha = .2, fill = "grey") + 
  annotate("rect", xmin = "1998", xmax = "2003", ymin = -Inf, ymax = Inf,
              alpha = .2, fill = "grey") + 
  annotate("rect", xmin = "2015", xmax = "2018", ymin = -Inf, ymax = Inf,
              alpha = .2, fill = "grey") + 
  facet_wrap(~series_name_rec, scales = "free")

```


-   Ugandan Bush War, **1980-1986**. [Wikipedia](https://en.wikipedia.org/wiki/Ugandan_Bush_War)
-   Second Congo War, "Operation Kitona" was a Rwandan/Ugandan offensive that marked the beginning of the Second Congo War **1998-2003**. [Wikipedia](https://en.wikipedia.org/wiki/Second_Congo_War)
-   Burundian unrest, **2015-2018**. [Wikipedia](https://en.wikipedia.org/wiki/Burundian_unrest_(2015%E2%80%932018))

---

#### 4. Debt service and Net Flows on external debt, general government sector (PPG)

Generally, the `debt service on external debt, general government sector (PPG)` has gradually been increasing as the debts pile up which is a good thing, however, we see that despite Kenya's efforts on `debt service on external debt, general government sector(PPG)` a huge chunk of that might be attributed to the high `Net flows on external debt, general government sector(PPG)` which are the payments that go the interests as opposed to the principal payments.

```{r, fig.height = 4, fig.width = 12.5, fig.align = "center"}
# we now wrangle our data

data2 %>%
  filter(country_name %in% nations) %>%
  filter(series_name %in% government_series) %>%
  group_by(series_name) %>%
  pivot_longer(cols = contains("_yr"), names_to = "year") %>%
  mutate(year = str_match(year, "\\d{4}")) %>%
  select(
    year,
    country_name,
    series_name,
    series_name_rec,
    value
  ) %>%
  mutate(value = as.numeric(na_if(value, ".."))) %>%
  replace(is.na(.), 0) %>%
  mutate(value = value / 10^6) %>%
  clean_names() %>%
   filter(series_name %in% c("Debt service on external debt, general government sector (PPG) (TDS, current US$)",
"Net flows on external debt, general government sector (PPG) (NFL, current US$)")) %>%
    ggplot(aes(year, value, color = country_name)) +
  geom_line(aes(group = country_name)) +
  scale_x_discrete(
        breaks = seq(1970, 2020, 2)) +
  theme(axis.text.x = element_text(angle = 90), 
        panel.background = element_rect(fill = "#F5E6B6")) +
  theme_hc() +
    scale_color_manual(
    values = c(Kenya = "#FF0000",
    Rwanda = "#046C9A",
    Tanzania = "#00A08A",
    Uganda = "#000000",
    Burundi = "#F98400")
  ) +
    labs(
    x = " ",
    y = "million USDs",
    caption = "- data from 1970-2020\n- each y-axis is independent",
    color = "COUNTRY"
  ) +
  facet_wrap(~series_name_rec, scales = "free")

```

---

#### 5. Principal debt forgiven and Principal debt rescheduled

No general trends visible on `Principal debt forgiven` and `Principal debt rescheduled`.


```{r, fig.height = 4, fig.width = 12.5, fig.align = "center"}

# we now wrangle our data
  
data2 %>%
  filter(country_name %in% nations) %>%
  filter(series_name %in% princ_other_series) %>%
  group_by(series_name) %>%
  pivot_longer(cols = contains("_yr"), names_to = "year") %>%
  mutate(year = str_match(year, "\\d{4}")) %>%
  select(
    year,
    country_name,
    series_name,
    series_name_rec,
    value
  ) %>%
  mutate(value = as.numeric(na_if(value, ".."))) %>%
  replace(is.na(.), 0) %>%
  mutate(value = value / 10^6) %>%
  clean_names() %>% 
   filter(series_name %in% c("Principal forgiven (current US$)", "Principal rescheduled (current US$)"
)) %>%
    ggplot(aes(year, value, color = country_name)) +
  geom_line(aes(group = country_name)) + 
  scale_x_discrete(
        breaks = seq(1970, 2020, 2)) +
  theme(axis.text.x = element_text(angle = 90)) +
  theme_hc() +
    scale_color_manual(
    values = c(Kenya = "#FF0000",
    Rwanda = "#046C9A",
    Tanzania = "#00A08A",
    Uganda = "#000000",
    Burundi = "#F98400")
  ) +
      labs(
    x = " ",
    y = "million USDs",
    caption = "- data from 1970-2020\n- each y-axis is independent",
    color = "COUNTRY"
  ) +
  facet_wrap(~series_name_rec, scales = "free")

```

---

#### 6. Use of IMF credits

Simply put, The IMF credits are low interest loans offered to countries in need. 
We see a sudden drop in reported in 2006 which might be attributed to policy changes as we see a pick up on the SDR allocations post-2008.

The IMF report linked below the plot might clarify things but the technical jargon was two vocabularies away from turning my brains into mush. `r emo::ji("laugh")`


```{r, fig.height = 4, fig.width = 12.5, fig.align = "center"}

# we now wrangle our data
  
data2 %>%
  filter(country_name %in% nations) %>%
  filter(series_name %in% imf_cred_series) %>%
  group_by(series_name) %>%
  pivot_longer(cols = contains("_yr"), names_to = "year") %>%
  mutate(year = str_match(year, "\\d{4}")) %>%
  select(
    year,
    country_name,
    series_name_rec,
    series_name,
    value
  ) %>%
  mutate(value = as.numeric(na_if(value, ".."))) %>%
  replace(is.na(.), 0) %>%
  mutate(value = value / 10^6) %>%
  clean_names() %>% 
    ggplot(aes(year, value, color = country_name)) +
  geom_line(aes(group = country_name)) + 
  scale_x_discrete(
        breaks = seq(1970, 2020, 2)) +
  theme(axis.text.x = element_text(angle = 90)) +
  theme_hc() +
    scale_color_manual(
    values = c(Kenya = "#FF0000",
    Rwanda = "#046C9A",
    Tanzania = "#00A08A",
    Uganda = "#000000",
    Burundi = "#F98400")
  ) +
        labs(
    x = " ",
    y = "million USDs",
    caption = "- data from 1970-2020\n- each y-axis is independent",
    color = "COUNTRY"
  ) + 
    annotate("rect", xmin = "2006", xmax = "2008", ymin = -Inf, ymax = Inf,
              alpha = .2, fill = "grey") + 
  facet_wrap(~series_name_rec, scales = "free")

```

- [download IMF annual report 2006](https://www.imf.org/~/media/Websites/IMF/imported-flagship-issues/external/pubs/ft/ar/2006/eng/pdf/_file11pdf.ashx)


---

### Conclusion

Despite several interesting findings, My main question about what happened in the 2000s has not been answered with the data at hand.

However, upon browsing I stumbled upon various resources that reported record high loans from our friends from the **far** east, China, from the year 2000. I specifically liked the dashboard on the following [site.](https://chinaafricaloandata.bu.edu) (click on explore) 

And with that I close this chapter with my eyes set on the public database provided by China Africa Research Initiative and Boston University Global Development Policy Center. 2021. It would be nice to explore in the near future.

---

## List of Events

This list is not an exhaustive list of all the major events that happened between 1970 - 2020

-   Breakup of East African Community, **1977**
-   Kagera War "Vita Vya Kagera", **1979**
-   General elections held in Uganda, **1980**. They followed the overthrow of Idi Amin the previous year and were the first since the pre-independence elections in 1962.
-   Ugandan Bush War, **1980-1986**. [Wikipedia](https://en.wikipedia.org/wiki/Ugandan_Bush_War)
-   End of Ujamaa in Tanzania, **1985**
-   KANU one-party rule ended in Kenya (Dec, **1991**) [report](https://www.hrw.org/reports/1992/WR92/AFW-03.htm)
-   Tanzania adopts Multiparty democracy system, **1992**. [report that speaks on the roles of donors in the democratization process.](https://media.africaportal.org/documents/rr7.pdf)
-   The World Bank, the International Monetary Fund (IMF) and other multilateral, bilateral and commercial creditors began the Heavily Indebted Poor Countries (HIPC) Initiative in **1996**, However among the East African countries, Kenya does not meet the criteria to receive debt relief. Countries must make full use of traditional debt-relief mechanisms to be eligible for debt relief under the HIPC Initiative. [Brief](https://www.worldbank.org/en/topic/debt/brief/hipc) [Countries must make full use of traditional debt-relief mechanisms to be eligible for debt relief under the HIPC Initiative.](https://www.imf.org/external/np/hipc/cost2/index.htm)
-   Second Congo War, "Operation Kitona" was a Rwandan/Ugandan offensive that marked the beginning of the Second Congo War **1998-2003**
- Jubilee **2000** [report](https://www.advocacyinternational.co.uk/featured-project/jubilee-2000)
-   almost steady increase in China to Africa loans from **2000**. [Report Link](https://chinaafricaloandata.bu.edu!)
-   SAPs underwent another transition, the introduction of Poverty Reduction Strategy Papers. **2002** [Doc](https://www.imf.org/external/np/prsp/prsp.aspx)
-   In Kenya, loans and grants from China became significant in size after **2002** when a new government was elected, when China's share in total aid exceeded 1 percent. Since then, China appears in Kenyan national statistics among bilateral donors whereas before then, it was classified in the category of 'other donors'. [report](https://www.brookings.edu/wp-content/uploads/2016/06/01_kenya_aid_mwega.pdf)
-   "Jubilee 2000", "Drop the Debt" 2005, "One" 2005, "Live 8", "Bono on the White House Lawn", "make poverty history" [blog](https://blogs.worldbank.org/africacan/african-debt-since-debt-relief-how-clean-is-the-slate)
-   keynote address by the president of Tanzania, Hon. Benjamin William Mkapa at annual conference of the jubilee debt campaign, London, **2005** ["WIPE OUT DEBT IN 2005"](https://www.tanzania.go.tz/egov_uploads/documents/WIPE_OUT_DEBT_IN_2005,,,_sw.pdf)
-   In June **2005**, the Group of Eight (G8) gathering of the worlds most developed economies, resolved, through their finance ministers, to grant total debt relief to 18 countries that had reached 'completion point', amounting to \$40 billion. Fourteen of these countries are in Africa. [Foreign aid, debt relief and Africa's development: problems and prospects. 2009.](https://www.tandfonline.com/doi/pdf/10.1080/10220460902986180)
-   [Tanzania Gets US\$ 75 Million Poverty Reduction Loan and US\$ 645 million Debt Relief, 2006](https://www.afdb.org/en/news-and-events/tanzania-gets-us-75-million-poverty-reduction-loan-and-us-645-million-debt-relief-4064)
-   Huge debt reliefs reported for Tanzania, **2006** [Tanzania Gets US\$ 75 Million Poverty Reduction Loan and US\$ 645 million Debt Relief, 2006](https://www.afdb.org/en/news-and-events/tanzania-gets-us-75-million-poverty-reduction-loan-and-us-645-million-debt-relief-4064)
-   [BOT's press release commemorating 50 years of independence, with annex explaining background to Tanzania's National debt, 2006](https://www.bot.go.tz/Adverts/PressRelease/sw/202003130724034239.pdf)
-   Drop of aid, **2006**[Development aid from OECD countries fell 5.1% in 2006](https://www.oecd.org/general/developmentaidfromoecdcountriesfell51in2006.htm)
-   Financial crisis, **2008**. [The 2007--2008 Financial Crisis in Review](https://www.investopedia.com/articles/economics/09/financial-crisis-review.asp)
-   East Africa Drought **2011**. [Said to be "the worst in 60 years" the drought caused a severe food crisis across Somalia, Djibouti, Ethiopia and Kenya that threatened the livelihood of 9.5 million people.](https://en.wikipedia.org/wiki/2011_East_Africa_drought)
-   Kenya peak debt, **2014** [Kenya Financial Sector Stability Report. 2014](https://www.centralbank.go.ke/uploads/financial_sector_stability/1205312556_Kenya%20Financial%20Sector%20Stability%20Report%202014.pdf)
-   Burundian unrest, **2015-2018**. [Wikipedia](https://en.wikipedia.org/wiki/Burundian_unrest_(2015%E2%80%932018))
-   [Central & East Africa Floods **2020**](https://appliedsciences.nasa.gov/what-we-do/disasters/disasters-activations/central-and-east-africa-floods-2020)

---